---
title: "dataRetrieval fpr USGS stream data"
subtitle: ""
author: "Mya Hardee"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    number-sections: true
editor_options: 
  chunk_output_type: inline
---

This script explains the use of dataretrival package to retrieve USGS stream data from the GAGES II dataset. This dataset was used in the 2025 Araki et al. paper.

```{r}
install.packages("dataRetrieval")
library(dataRetrieval)
```

Here is the link with information on using the package: https://doi-usgs.github.io/dataRetrieval/


##Here, I will pull dicrete data from the hickory creek BasinID, located in the basinID file: https://www.sciencebase.gov/catalog/item/59692a64e4b0d1f9f05fbd39

#This is specifically pH data.

```{r}
hickory_discdata_ph <- read_waterdata_samples(
  monitoringLocationIdentifier = "USGS-08052780",
  characteristicUserSupplied = "pH, water, unfiltered, field",
  activityStartDateUpper = "2000-01-01",
  dataProfile = "narrow")

  ## bounding box function (if used) must be: west, south, east, north
```

To find codes for other variables for characteristics: https://waterdata.usgs.gov/code-dictionary/

##OTHER TYPES OF DATA TO PULL

Other than discrete data, you can pull other types of data from the monitors using different code.

The example below is pulled from the instructional page for USGS daily data. The parameter code indicates discharge in ft cubed/s

```{r}
site <- "USGS-02238500"
pcode <- "00060"
dv_data_sf <- read_waterdata_daily(monitoring_location_id = site,
                              parameter_code = "00060", 
                              time = c("2021-01-01", "2022-01-01"))
```
#Here is one for hickory

```{r}
dv_data_sf <- read_waterdata_daily(monitoring_location_id = "USGS-08052780",
                              parameter_code = "00060", 
                              time = c("2021-01-01", "2022-01-01"))

```

## Overall, this can be a good source for pulling this data from this pre-existing monitor. 